Manuscript_draft

### 2021-03-16 manuscript_draft.Rmd
remove 1992 data from analysis

  # Create TU lookup
  mutate(BA.Target=paste0("TU:",BA.Target))

@@

To test the hypothesis that tree-level radial growth response over the 1992-2019 period varied with interactions between (i) tree diameter at the time of treatment (1992); (ii) treatment; and (iii) species, we fit linear mixed effects models with log-transformed 1992-2019 diameter increment as the response variable, and treatment, species and log-transformed 1992 diameter as interacting fixed effects. To account for plot-level productivity differences, we included plot as a random intercept effect.  Models were fit using the 'lme4' package for R [@].  We also fit a similar model with the 1992-2019 diameter increment expressed as a percentage of 1992 diameter.  Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.  Variance explained for each model are reported using the approach described in Nakagawa et al. [-@Nakagawa2013], as implemented in the MuMIn package for R [@barton2009mumin].  Pairwise contrasts were performed using the 'emmeans' package for R [@] and Tukey HSD method of *p* value adjustment was used to control for multiplicity.  We report all significant results at p<0.05.

@@

### Basal area increment from 1992-2019

@@

# Filter only for trees that have measurements in 1992 and 2019 
treeList<- # assign variable
  X %>% 
  filter(TreeID!="New") %>% # remove new trees
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(Species %in% c("Spruce","Fir")) %>% # filter birch and fdi out of dataset
  filter(Status=="Live") %>% # filter for live trees
  filter(Year %in% c(1992,2019)) %>% #filter for start and end dates
  group_by(TreeID) %>% 
  summarise(N=n()) %>% 
  filter(N==2) %>% #filter for trees with 5 measurements
  pull(TreeID)

# derive periodic increment (period diameter or basal area increment)
piDat<-
  X %>% 
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(TreeID %in% treeList) %>% 
  mutate(BA=d2ba(DBH)) %>% # convert DBH to basal area
  dplyr::select(Plot,TreeID,Species,Year,BA) %>% 
  pivot_wider(names_from=Year,values_from=BA) %>% 
  mutate(PI=`2019`-`1992`) %>% # 
  filter(PI>0) %>% # one tree has a PI of 0, has to be removed
  mutate(PIpct=PI/`1992`*100) %>% # express periodic increment a a percentage
  dplyr::select(Plot,TreeID,Species,PBAI=PI,PBAIpct=PIpct,Init=`1992`) %>% 
  left_join(plotDat,by="Plot")

@@

  pairs(rg,by=c("BA.Target","Init"),type="response") %>% 
  as.data.frame() %>%
    mutate(Init=as.numeric(as.character(Init))) %>% # change from factor to numeric
    mutate(Init=ba2d(Init)) %>% # convert basal area to diameter
  arrange(BA.Target,Init) %>% 
  rename(DBH.1992="Init")

@@

trtContrast<-
  pairs(rg,by=c("Species","Init"),type="response") %>% 
  as.data.frame() %>% 
  mutate(Init=as.numeric(as.character(Init))) %>% # change from factor to numeric
  mutate(Init=ba2d(Init)) %>% # convert basal area to diameter
  arrange(Species,contrast,Init) %>% 
  rename(DBH.1992="Init")

@@

The tree-level basal area increment model indicated that interactions between log-transformed tree size, species and treatment explained `r round(r.squaredGLMM(fitBA)[2]*100,1)`% of the variation in log-transformed 1992-2019 tree-level BAI, and `r round(r.squaredGLMM(fitBApct)[2]*100,1)`% of the variation in basal area increment as a percentage of tree size in 1992 (p<0.0001 for both, `r tableNums("fitBA.table",display="cite")`).  Estimated marginal means from the model showed that between species and among treatments, basal area increment increased with tree size, meaning that larger trees increased basal area more than smaller trees (`r figNums("treeGrowth.plot",display="cite")`), although smaller trees increased their size proportionally more than large trees (`r figNums("treeGrowthPct.plot",display="cite")`).  Species contrasts showed that spruce trees increased  basal area significantly more than than fir in the high-removal and control treatment units, whereas basal growth between the two species was close to equal in the low-removal unit (`r appNums("sppContrast.table",display="cite")`).  Within-species contrasts showed that spruce basal area increment in the high-removal unit exceeded spruce growth in both other units (`r appNums("trtContrast.table",display="cite")`), whereas fir basal area increment in the high- and low-removal units was higher than the control unit, and there were no significant differences between the two harvested units (`r appNums("trtContrast.table",display="cite")`).

@@

  # Determine which plots have measurements in 1992 and 2019
  X %>% 
   filter(Year %in% c(1992,2019)) %>% 
   filter(Status=="Live") %>%

@@

  # Filter for years of interest
  filter(Year %in% c(1992,2019)) %>% 
  mutate(Year=fct_drop(Year)) %>% 
  filter(Status=="Live") %>%

@@

plotData.X<-
# Determine which plots have measurements in 1992 and 2019
  X %>%

@@

perInc.X<-
  plotData.X %>% 
  pivot_longer(cols=Vol.sum:QMD.mean,names_to="var",values_to = "val") %>% 
  pivot_wider(names_from = Year,values_from = val) %>% 
  mutate(Period1=(`1994`-`1992`)/3,
         Period2=(`1997`-`1994`)/3,
         Period3=(`2009`-`1997`)/11,
         Period4=(`2019`-`2009`)/11) %>% 
  dplyr::select(BA.Target:var,Period1:Period4) %>% 
  pivot_longer(cols=Period1:Period4,names_to="Period",values_to="Values") %>% 
  pivot_wider(names_from=var,values_from=Values) %>%

@@

  # rename periods
  mutate(Period=dplyr::recode(Period,
                       Period1="1992-1994",
                       Period2="1994-1997",
                       Period3="1997-2009",
                       Period4="2009-2019"))

@@

# Create slope contrasts table
standFit.contrasts<-
  data.frame(Variable="Volume",vol.contrast) %>% 
  rbind(data.frame(Variable="Basal area",ba.contrast)) %>% 
  rbind(data.frame(Variable="Stand density",sph.contrast)) %>% 
  filter(Year_revpairwise %in% c("2019 - 1992")) %>% 
  dplyr::select(Variable,`Year contrast`="Year_revpairwise",`Treatment contrast`="BA.Target_pairwise",everything()) %>% 
  rename(`t ratio`="t.ratio",
         `p value`="p.value") %>% 
  mutate_at(vars(estimate:`t ratio`),round,3) %>% 
  mutate_at(vars(`p value`),round,5)

@@

# Create species mean contrasts table

@@

### Rate of change from 1992 to 2019
All treatment units, including the control, increased volume, basal area and stand density over the 1992-2019 period (`r figNums("standPlots",display="cite")` and `r appNums("standPlot.table",display="cite")`).  Mean annual basal area and volume increment peaked during the 1997-2009 period in the two harvested treatments, and peaked in the 1994-1997 period in the control treatment unit (`r figNums("perInc.plot",display="cite")`).  From 1992 to 2019, estimated marginal means of basal area increment in the low RBA unit exceeded marginal means in the high RBA and control unit by an estimated 
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[1]`
and
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[2]`
m² per hectare, respectively (p<0.01 for both, `r appNums("standFit.contrasts",display="cite")`).

@@

@@

    # Cumulative stand basal area
    mutate(Cumu.BA=cumsum(BA.Init)) %>% 
    mutate(Cumu.BA2=max(Cumu.BA)) %>% 
    mutate(Cumu.Prop.BA=Cumu.BA/Cumu.BA2) %>% 
    dplyr::select(-Cumu.BA2)

    # Calculate Growth Dominance statistic
    gd.stats<-
      gd.cumu.Inc %>% 
      mutate(GDS=(Cumu.Prop.Inc+lag(Cumu.Prop.Inc,default=0))*(Cumu.Prop.BA-lag(Cumu.Prop.BA,default=0))) %>% 
      ungroup() %>% 
      group_by(Plot,Year) %>% 
      summarise(GDS=1-sum(GDS))

    gd.stats %>% 
      ggplot()+
      aes(x=Year,y=GDS)+
      geom_point()+
      facet_wrap(~Plot)

    ggplot()+
    aes(x=Cumu.Prop.BA,y=Cumu.Prop.Inc)+
    geom_point()+
    geom_line()+
    geom_abline(slope=1,intercept=0,linetype = "dashed")+
    theme(aspect.ratio = 1)+
    facet_wrap(~Year,scales="free_y")

@@

>>>>

  # Create TU lookup
  mutate(BA.Target=paste0("TU:",BA.Target)) %>%

  # need to remove plots 16-19 because they were only measured once
  filter(!Plot %in% c(16,17,18,19))

@@

To test the hypothesis that tree-level radial growth response over the 1994-2019 period varied with interactions between (i) tree diameter at the time of final treatment (1994); (ii) treatment; and (iii) species, we fit linear mixed effects models with log-transformed 1994-2019 diameter increment as the response variable, and treatment, species and log-transformed 1992 diameter as interacting fixed effects. To account for plot-level productivity differences, we included plot as a random intercept effect.  Models were fit using the 'lme4' package for R [@].  We also fit a similar model with the 1994-2019 diameter increment expressed as a percentage of 1994 diameter.  Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.  Variance explained for each model are reported using the approach described in Nakagawa et al. [-@Nakagawa2013], as implemented in the MuMIn package for R [@barton2009mumin].  Pairwise contrasts were performed using the 'emmeans' package for R [@] and Tukey HSD method of *p* value adjustment was used to control for multiplicity.  We report all significant results at p<0.05.

@@

### Basal area increment from 1994-2019

@@

# Filter only for trees that have measurements in 1992 and 2019 
treeList<- # assign variable
  X %>% 
  filter(TreeID!="New") %>% # remove new trees
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(Species %in% c("Spruce","Fir")) %>% # filter birch and fdi out of dataset
  filter(Status=="Live") %>% # filter for live trees
  filter(Year %in% c(1994,2019)) %>% #filter for start and end dates
  group_by(TreeID) %>% 
  summarise(N=n()) %>% 
  filter(N==2) %>% #filter for trees with 5 measurements
  pull(TreeID)

# derive periodic increment (period diameter or basal area increment)
piDat<-
  X %>% 
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(TreeID %in% treeList) %>% 
  mutate(BA=d2ba(DBH)) %>% # convert DBH to basal area
  dplyr::select(Plot,TreeID,Species,Year,BA) %>% 
  pivot_wider(names_from=Year,values_from=BA) %>% 
  mutate(PI=`2019`-`1994`) %>% # 
  filter(PI>0) %>% # one tree has a PI of 0, has to be removed
  mutate(PIpct=PI/`1994`*100) %>% # express periodic increment a a percentage
  dplyr::select(Plot,TreeID,Species,PBAI=PI,PBAIpct=PIpct,Init=`1994`) %>% 
  left_join(plotDat,by="Plot")

@@

  pairs(rg,by=c("BA.Target","Init"),type="response") %>% 
  as.data.frame() %>%
    mutate(Init=as.numeric(as.character(Init))) %>% # change from factor to numeric
    mutate(Init=ba2d(Init)) %>% # convert basal area to diameter
  arrange(BA.Target,Init) %>% 
  rename(DBH.1994="Init")

@@

trtContrast<-
  pairs(rg,by=c("Species","Init"),type="response") %>% 
  as.data.frame() %>% 
  mutate(Init=as.numeric(as.character(Init))) %>% # change from factor to numeric
  mutate(Init=ba2d(Init)) %>% # convert basal area to diameter
  arrange(Species,contrast,Init) %>% 
  rename(DBH.1994="Init")

@@

The tree-level basal area increment model indicated that interactions between log-transformed tree size, species and treatment explained `r round(r.squaredGLMM(fitBA)[2]*100,1)`% of the variation in log-transformed 1994-2019 tree-level BAI, and `r round(r.squaredGLMM(fitBApct)[2]*100,1)`% of the variation in basal area increment as a percentage of tree size in 1994 (p<0.0001 for both, `r tableNums("fitBA.table",display="cite")`).  Estimated marginal means from the model showed that between species and among treatments, basal area increment increased with tree size, meaning that larger trees increased basal area more than smaller trees (`r figNums("treeGrowth.plot",display="cite")`), although smaller trees increased their size proportionally more than large trees (`r figNums("treeGrowthPct.plot",display="cite")`).  Species contrasts showed that spruce trees increased  basal area significantly more than than fir in the high-removal and control treatment units, whereas basal growth between the two species was close to equal in the low-removal unit (`r appNums("sppContrast.table",display="cite")`).  Within-species contrasts showed that spruce basal area increment in the high-removal unit exceeded spruce growth in both other units (`r appNums("trtContrast.table",display="cite")`), whereas fir basal area increment in the high- and low-removal units was higher than the control unit, and there were no significant differences between the two harvested units (`r appNums("trtContrast.table",display="cite")`).

@@

  # Determine which plots have measurements in 1992 and 2019
  X %>% 
   filter(Year %in% c(1994,2019)) %>% 
   filter(Status=="Live") %>%

@@

  # Filter for years of interest
  filter(Year %in% c(1994,2019)) %>% 
  mutate(Year=fct_drop(Year)) %>% 
  filter(Status=="Live") %>%

@@

plotData.X<-

  X %>%

@@

perInc.X<-
  plotData.X %>% 
  pivot_longer(cols=Vol.sum:QMD.mean,names_to="var",values_to = "val") %>% 
  pivot_wider(names_from = Year,values_from = val) %>% 
  mutate(Period1=(`1997`-`1994`)/3,
         Period2=(`2009`-`1997`)/11,
         Period3=(`2019`-`2009`)/11) %>% 
  dplyr::select(BA.Target:var,Period1:Period3) %>% 
  pivot_longer(cols=Period1:Period3,names_to="Period",values_to="Values") %>% 
  pivot_wider(names_from=var,values_from=Values) %>%

@@

  # rename periods
  mutate(Period=dplyr::recode(Period,
                       Period1="1994-1997",
                       Period2="1997-2009",
                       Period3="2009-2019"))

@@

# Create slope contrasts table
standFit.contrasts<-
  data.frame(Variable="Volume",vol.contrast) %>% 
  rbind(data.frame(Variable="Basal area",ba.contrast)) %>% 
  rbind(data.frame(Variable="Stand density",sph.contrast)) %>% 
  filter(Year_revpairwise %in% c("2019 - 1994")) %>% 
  dplyr::select(Variable,`Year contrast`="Year_revpairwise",`Treatment contrast`="BA.Target_pairwise",everything()) %>% 
  rename(`t ratio`="t.ratio",
         `p value`="p.value") %>% 
  mutate_at(vars(estimate:`t ratio`),round,3) %>% 
  mutate_at(vars(`p value`),round,5)

@@

@@

### Rate of change from 1994 to 2019
All treatment units, including the control, increased volume, basal area and stand density over the 1994-2019 period (`r figNums("standPlots",display="cite")` and `r appNums("standPlot.table",display="cite")`).  Mean annual basal area and volume increment peaked during the 1997-2009 period in the two harvested treatments, and peaked in the 1994-1997 period in the control treatment unit (`r figNums("perInc.plot",display="cite")`).  From 1994 to 2019, estimated marginal means of basal area increment in the low RBA unit exceeded marginal means in the high RBA and control unit by an estimated 
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[1]`
and
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[2]`
m² per hectare, respectively (p<0.01 for both, `r appNums("standFit.contrasts",display="cite")`).

@@

# # sgr.Dat
#   sgr.Dat<-
#     gd.Dat %>%
#     filter(Var=="PBAI") %>% 
#     ungroup() %>% 
#     group_by(Plot) %>% 
#     mutate(Stand.Inc=sum(Inc),
#            Stand.BA=sum(BA.Init)) %>% 
#     mutate(Prop.growth=Inc/Stand.Inc,
#            Prop.size=BA.Init/Stand.BA) %>% 
#     left_join(plotDat,by="Plot")

@@

    # Cumulative stand basal area
    mutate(Cumu.BA=cumsum(BA.Init)) %>% 
    mutate(Cumu.BA2=max(Cumu.BA)) %>% 
    mutate(Cumu.Prop.BA=Cumu.BA/Cumu.BA2) %>% 
    dplyr::select(-Cumu.BA2) %>%

    # Bring in plot data
     left_join(plotDat,by="Plot")

    # Calculate Growth Dominance statistic
    gd.stats<-
      gd.cumu.Inc %>% 
      mutate(GDS=(Cumu.Prop.Inc+lag(Cumu.Prop.Inc,default=0))*(Cumu.Prop.BA-lag(Cumu.Prop.BA,default=0))) %>% 
      ungroup() %>% 
      group_by(Plot,Year) %>% 
      summarise(GDS=1-sum(GDS)) %>% 
      left_join(plotDat,by="Plot") %>% 
      ungroup() %>% 
      group_by(BA.Target,Year) %>% 
      summarise(GDS.mean=round(mean(GDS),3),
                GDS.sd=round(sd(GDS),2)) %>% 
      mutate(GDS.label=paste0("GDS: ",GDS.mean," (",GDS.sd,") "))

  #   gd.stats.plot<-
  #     gd.stats %>% 
  #   filter(Year %in% c(2009,2019)) %>% 
  # ggplot()+
  # aes(x=BA.Target,y=GDS,fill=Year)+
  # geom_boxplot()
  #

  # Plotting
  # ## Growth Dominance statistic
  gd.plot<-
    gd.cumu.Inc %>%
    filter(Year %in% c(2009,2019)) %>%
    ggplot()+
    aes(x=Cumu.Prop.BA,y=Cumu.Prop.Inc,group=BA.Target)+
    geom_point(colour="grey")+
    geom_smooth(se=F,size=1.5)+
    geom_abline(slope=1,intercept=0,linetype = "dashed",size=1.5)+
    theme(aspect.ratio = 1,
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
    facet_grid(Year~BA.Target,scales="free_y")+
    xlab("Cumulative proportion of stand size (basal area)")+
    ylab("Cumulative proportion of stand growth (basal area)")+
    facet_grid(Year~BA.Target)+
    geom_text(data=filter(gd.stats %>% filter(Year %in% c(2009,2019))),
            size=4,
            mapping = aes(x = Inf, y = 0.1, label = GDS.label),
            hjust="right")

  ## Plot measurement year GSD
      # gd.stats %>% 
      #   ggplot()+
      #   aes(x=Year,y=GDS,group=Plot,color=Plot)+
      #   geom_line()+
      #   facet_wrap(~BA.Target)
      #

  ## GSize-Growth relationship statistic
      # sgr.ind.plot<-
      #   sgr.Dat %>% 
      #  ggplot()+
      #   aes(x=Prop.size,y=Prop.growth,group=Plot,color=BA.Target)+
      #   geom_point()+
      #   geom_smooth(aes(group=Plot),method="lm",se=F,color="black")+
      #   facet_wrap(~Year)+
      #       theme(aspect.ratio = 1)+
      #     facet_grid(BA.Target~Year)

@@

Across all plots and time periods, smaller trees contributed proportionally more to stand basal area growth than larger trees (`r figNums("gds.plot",display="cite")`).  Over the last two measurement periods, the growth dominance statistic decreased in all plots except one (not shown), indicating that individual tree growth contributions to stand growth became more proportional to tree size over the last 11 growing seasons.
